\

```{r}
library(dplyr)
library(ggplot2)

filter(starwars, species == "Human")

filter(starwars, species == "Human", sex == "male")
```

We can use different comparison operators to filter:

-   ! is "not"

-   == is "equal"

-   \$ is "and"

-   \| is "or"

-   \< is "less than"

-   \> is "greater than"

-   is.na() is

```{r}
## Returns characters with a missing mass
filter(starwars, is.na(mass))
```

```{r}
## How to get all characters that are not human?
filter(starwars, species != "Human")
```

```{r}
skin_color <- "grey"
skin_color %in% c("yellow", "fair", "tan")

## How would we filter for characters with brown or blond hair
filter(starwars, hair_color %in% c("blond", "brown"))

## hair_color %in% c("blond", "brown") is checking if the hair_color value for each row in the starwars dataframe is either "blond" or "brown". So, the full expression filter(starwars, hair_color %in% c("blond", "brown")) will return a subset of the starwars dataframe where the hair_color is either "blond" or "brown". %in% Returns that are eiether blond or brown


```

grepl("characters_of_interest", variable_name) returns True if the value of the variable_name contains the characters_of_interest.

```{r}
filter(starwars, grepl("blond", hair_color))


## How would we get all characters with blue in their eyes
filter(starwars, grepl("blue", eye_color))
```

you can use `<`, `>`, `<=`, `>=` to compare numeric variables or categorical variables

-   nominal categorical variables(character vectors) are ranked alphabetically

-   ordinal categorical variables(factors) are ranked based on the built-in order

```{r}
## starwars characters with a name starting with letters before "D"
filter(starwars, name < "D") # ranking is based on alphabetical order

str(diamonds)
filter(diamonds, cut < "Good") # ranking is based on levels of the ordered factor cut

"D" < "d" # false
"d" > "D" # true 
```

`between(x, left, right)` is a shorthand for `x >= left` & `x <= right`

```{r}
between(5,1,7) # is 5 between 1 and 7? True because it 5 is less than 7 and greater than 1
filter(starwars, between(mass, 50, 100)) # mass values between 50 and 100
```

### Practice!

Find all star wars characters that:

1.  Are missing a hair color

2.  Are shorter than 100 cm and are not droids

3.  Were born between the years 100 and 200

4.  Anything of my choice

```{r}
# 1
filter(starwars, is.na(hair_color))
```

```{r}
# 2
filter(starwars, height < 100, species != "Droid")
```

```{r}
# 3
filter(starwars, between(birth_year, 100, 200))
```

```{r}
#4 
filter(starwars, hair_color == "white")
```

### Order Rows with `Arrange`

```{r}
arrange(starwars, name)
## order by aAbBcCdD

arrange(starwars, species, birth_year)

arrange(starwars, species, desc(birth_year))
```

```{r}
arrange(starwars, desc(height))


arrange(starwars, height, species)
arrange(starwars, species, height)
```

### Select Function

```{r}
starwars
select(starwars, -films, -vehicles, -starships)
```

Helpful functions

everything() is a shortcut for "everything else"

```{r}
# Re-order columns, species goes in front
select(starwars, "species", everything()) # put species first, then all the other columns

## Alphabetize columns
starwars_names <- sort(names(starwars))
select(starwars,starwars_names)

```

stars_with("abc") matches column names that start with "abc"

ends_with("abc") matches column names that end with "abc"

containts("abc") matches column names that contain "abc"

```{r}
# select a column that starts with letter h, you can also do contains and ends with
select(starwars, starts_with("h"))
```

Practice!

1.  Select each character's homeworld and species. What question might this subset of variables help answer?

2.  Brainstorm as many ways as possible to select hair_color, skin_color, eye_color, and birth_year.

```{r}
# 1 

select(starwars, contains("name"), contains("homeworld"), contains("species"))

# This data set can tell us which species usually occur in which homeland more often, for example we can see that a lot of human species live on tatooine.
```

```{r}
# 2

select(starwars, contains("hair_color"), contains("skin"), contains("eye"), contains("birth"))
```

### `Mutate`

```{r}
starwars_small <- select(starwars, height, mass, species)
starwars_small
mutate(starwars_small, height_m = height/100)
```

We can also create multiple variables at once and we can refer to variables that we just created

```{r}
mutate(starwars_small,
       height_m = height/100,
       bmi = mass/height_m^2)

transmute(starwars_small,
       height_m = height/100,
       bmi = mass/height_m^2)
```

Summary statistics give one value for the entire column: sum(), mean()

```{r}
mutate(starwars_small,
       sum = sum(mass, na.rm = T))

mutate(starwars_small,
       row_addition = height + mass,
       summary_sum = sum(height + mass, na.rm = T))
```

Group by a variable using the .by parameter

```{r}
mutate(starwars_small,
       species_mass = sum(mass, na.rm = T),
       .by = species)
```

### 

### Practice!

1.  convert mass to pounds

```{r}
mutate(starwars_small, 
       converted_mass = mass*2.2)
```

### Summarize data with `Summarize`

```{r}
summarize(starwars,
          mean_mass = mean(mass, na.rm = T))

summarize(starwars,
          mean_mass = mean(mass, na.rm = T),
          count = n(),
          .by = species)
```
